Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-07-01 co-regular-EPQ-arxiv.tex
modify

\begin{lemma}
There exists a $\Gamma$-complete automaton for each $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$, $\mathsf{MatchedMaxPriority}\}$.
\end{lemma}

When defining $\Gamma$-complete automata, we assume that every implementation $\mathcal{I}$ behaves correctly, i.e., as a FIFO queue, when only values with the same priority are observed. More precisely, we assume that for every execution $e\in\mathcal{I}$ and every priority $p\in\mathbb{P}$, the projection of $e$ to values with priority $p$ is linearizable (w.r.t. $\seqPQ$). {\color {blue}This property can be checked separately using the techniques introduced for regular FIFO queues in~\cite{DBLP:conf/icalp/BouajjaniEEH15}} {\color {red}This property can be checked separately using register automata obtained from the finite automata in~\cite{DBLP:conf/icalp/BouajjaniEEH15} for FIFO queue} (see Appendix~\ref{sec:appendix lemma and register automata for FIFO of single-priority executions} for more details). Note that this assumption excludes some obvious violations, such as a $\textit{rm}(a)$ operation happens before a $\textit{put}(a,p)$ operation, for some $p$.

@@

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Let $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$ and $e$ a data-differentiated execution. Then, $e$ is $\Gamma$-linearizable iff $e\vert_{\preceq p}$ is $\Gamma$-linearizable for some maximal priority $p$ in $e$.
\end{restatable}
\begin {proof} (Sketch)
%{\color {blue}We prove an equivalent property: $e$ is $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is $\Gamma$-linearizable for some maximal priority $p$ in $e$.} %We consider the case of $\mathsf{MatchedMaxPriority}$ and the case of $\mathsf{UnmatchedMaxPriority}$ is similar.
To prove the $\textit{only if}$ direction, let $e$ be a data-differentiated execution linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$. Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$ imposes no restriction on the operations in $u$, $v$ and $w$ with priorities incomparable to $p$, erasing all these operations results in a sequential execution which still satisfies this property. Similarly, for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that is not $\mathsf{MatchedMaxPriority}^{>}$-linearizable. We represent each operation as a time interval whose left, resp., right, bound corresponds to the call, resp., return action. {\color {blue}Operations adding and removing the same value are aligned vertically.}}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

@@

The execution in \figurename~\ref{fig:introduce gap for EPQ1Lar} is not $\mathsf{MatchedMaxPriority}^>$-linearizable because the left-right constraint of the maximal priority value $b$ contains a cycle: $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$. The presence of such a cycle is equivalent to the execution not being $\mathsf{MatchedMaxPriority}^>$-linearizable (see Appendix~\ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}), {\color {red}as indicated by the following lemma:

\begin{restatable}{lemma}{LinEqualsConstraintforEPQOneLar}
\label{lemma:Lin Equals Constraint for EPQ1Lar}
Given a data-differentiated execution $e$ where $\mathsf{Has\text{-}MatchedMaxPriority}(e)$ holds, let $p$ be its maximal priority and $\textit{put}(x,p),\textit{rm}(x)$ are only operations of priority $p$ in $e$. Let $G$ be the graph representing the left-right constraint of $x$. $e$ is $\mathsf{MatchedMaxPriority}$-linearizable, if and only if $G$ has no cycle going through $x$.
\end{restatable}
}

@@

%TODO WHY IT IS A SET OF AUTOMATA ? SAY WHAT ARE THE CASES THAT NEED TO BE COVERED. SAY WHAT IS THE CASE WE GIVE IN THE FIGURE. USE $\top$ AS MENTIONED AT THE BEGINNING OF THE SECTION INSTEAD OF 	$c$. USE BIG CAPS FOR SETS OF TRANSITION LABELS.
To define a $\mathsf{MatchedMaxPriority}^>$-complete automaton we need to consider several cases depending on the order between the call/return actions of the $\textit{put}$/$\textit{rm}$ operations that add and respectively, remove the value $b$. For example, the case where the put happens-before the remove (as in \figurename~\ref{fig:introduce gap for EPQ1Lar}) is pictured in \figurename~\ref{fig:automata APQ1Lar-1 in paper}. {\color {red}This automaton captures the three possible ways of ordering the first action $\textit{ret}(\textit{put},a,\_)$ w.r.t. the actions with value $b$, which are pictured in \figurename~\ref{fig:executions APQ1Lar-1 in paper} (a)} (this action cannot occur after $\textit{call}(\textit{rm},b,\_)$ since $b$ must be covered by the $a$-s). The paths corresponding to these three possible orders are: $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_7$, $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_{10}$ and $q_1 \rightarrow q_9 \rightarrow q_{10} \ldots \rightarrow q_7$. {\color {red}In \figurename~\ref{fig:executions APQ1Lar-1 in paper}, we show all the four orders of the call/return actions of adding and removing $b$, and also possible orders of the first $\textit{ret}(\textit{put},a,\_)$ w.r.t the actions with value $b$. In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, three register automata is constructed according to the cases of \figurename~\ref{fig:executions APQ1Lar-1 in paper} (b), (c) and (d), respectively.}
{\color {blue}Automata for the other cases can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}.}

@@

\begin{figure}[htbp]
  \centering
  %\includegraphics[width=1 \textwidth]{figures/PIC_HIS_PQ1Lar-pprr-in-paper.pdf}
  \includegraphics[width=1 \textwidth]{figures/PIC_HIS_PQ1Lar-fouCase.pdf}
%\vspace{-10pt}
  %\caption{Three possible positions of the first $\textit{call}(\textit{put},a,\_)$ in captured execution}
  \caption{{\color {red}Four cases of ordering actions with value $b$, and possible orders of the first $\textit{ret}(\textit{put},a,\_)$ w.r.t the actions with value $b$}}
  \label{fig:executions APQ1Lar-1 in paper}
\end{figure}

@@

To recognize the scenarios in \figurename~\ref{fig:introduce pb order}, we introduce an order $<_{\textit{pb}}$ between values which intuitively, can be thought of as ``a value $a$ is {\color {red}put} before another value $b$''. %Let us introduce put-before order $<_{\textit{pb}}$ to formally state that ``an value is putted before another value''.
Thus, given a data-differentiated execution $e$ and two values $a$ and $b$ of maximal priority, $a <_{\textit{pb}} b$ if one of the following holds: (1) $\textit{put}(a,\_) <_{\textit{hb}} \textit{put}(b,\_)$, (2) $\textit{rm}(a) <_{\textit{hb}} \textit{rm}(b)$, or (3) $\textit{rm}(a) <_{\textit{hb}} \textit{put}(b,\_)$. Sometimes we use $a <_{\textit{pb}}^A b$, $a <_{\textit{pb}}^B b$ and $a <_{\textit{pb}}^C b$ to explicitly distinguish between these three cases. Let $<_{\textit{pb}}^*$ be the transitive closure of $<_{\textit{pb}}$.
%Intuitively, $a <_{\textit{pb}}^* b$ means that we can infer that $a$ is putted before $b$ with the help of several intermediate values.

Then, to model the time intervals in which a remove operation like $\textit{rm}(b)$ in \figurename~\ref{fig:introduce pb order}, can be linearized (outside of intervals of smaller priority values) we introduce the notion of gap-point. {\color {blue}This notion relies on a standard indexing of actions in an execution, starting with 0.} {\color {red}Here we assume that the index of actions of an execution starts from $0$.}

@@

\begin{restatable}{lemma}{EPQOneEqualAsPBandGP}
\label{lemma:EPQ1Equal as pb order and gap-point}
Let $e$ be a data-differentiated execution which contains only one maximal priority $p$ such that $\mathsf{Has\text{-}MatchedMaxPriority}(e)$ holds.
%Given a data-differentiated $p$-execution $e$ where $\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Condition}(e)$ holds.
Then, $e$ is not linearizable w.r.t $\mathsf{MatchedMaxPriority}^{=}$ iff $e$ contains two values $x$ and $y$ of maximal priority $p$ such that $y <_{\textit{pb}}^* x$, {\color {red}and the rightmost gap-point of $x$ is strictly smaller than the index of $\textit{call}(\textit{put},y,p)$ or $\textit{call}(\textit{rm},y)$}.
\end{restatable}

@@

\smallskip
To characterize violations to linearizability w.r.t. $\mathsf{MatchedMaxPriority}^{=}$, one has to consider all the possible orders between call/return actions of the operations on values $a$, $b$, and $a_i$ in Lemma \ref{lemma:ob order has bounded length}, and the right-most gap point of $b$. Excluding the inconsistent cases, {\color {blue}we are left with 5 possible orders that are listed in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}}{\color {red}we are left with 5 possible orders that are shown in \figurename~\ref{fig:five enumerations}, where $o$ denotes the rightmost gap-point of $b$}.
For each case, we define an automaton recognizing the induced set of violations. {\color {blue}For instance, one such case and the corresponding register automaton is shown in \figurename~\ref{fig:an enumeration and its witness automaton} ($o$ denotes the rightmost gap-point of $b$).} {\color {red}For instance, the register automata for the case of \figurename~\ref{fig:five enumerations}(a) is shown in \figurename~\ref{fig:an enumeration and its witness automaton}.} In this case, the conditions in Lemma~\ref{lemma:EPQ1Equal as pb order and gap-point} are equivalent to the fact that intuitively, the time interval from $\textit{call}(\textit{rm},a)$ to $\textit{ret}(\textit{rm},b)$ is covered by lower priority values (and thus, there is no gap-point of $b$ which occurs after $\textit{call}(\textit{rm},a)$). Using again the data-independence property, these lower priority values can all be renamed to a fixed value $d$, and the other values to a fixed value $\top$.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8 \textwidth]{figures/PIC-HIS-FiveEnumerations.pdf}
%\vspace{-10pt}
  \caption{{\color {red} Five cases that need to be considered.}}
  \label{fig:five enumerations}
\end{figure}

@@

We describe a class $\mathcal{C}$ of data-independent implementations for which linearizability w.r.t. $\seqPQ$ is decidable. The implementations in $\mathcal{C}$ allow an unbounded number of values but a bounded number of priorities. Each method manipulates a finite number of local variables which store Boolean values, or data values from $\mathbb{D}$. Methods communicate through a finite number of shared variables that also store Boolean values, or data values from $\mathbb{D}$. Data values may be assigned, but never used in program predicates (e.g., in the conditions of if-then-else and while statements) so as to ensure data independence. This class captures typical implementations, or finite-state abstractions thereof, e.g., obtained via predicate abstraction. {\color{blue}Since the $\Gamma$-complete automata $A(\Gamma)$ we define use}{\color {red}Since the $\Gamma$-complete automata $A(\Gamma)$ uses} a fixed set {\color {red}$D=\{a,b,a_1,d,e,\top\}$ of values}, we have that {\color {red}$\mathcal{C}\cap A(\Gamma)\neq\emptyset$ for some $\Gamma$ iff $\mathcal{C}_D\cap A(\Gamma)\neq\emptyset$ where $\mathcal{C}_D$ is the subset of $\mathcal{C}$ that uses only values in $D$.} {\color{blue}The set of executions $\mathcal{C}_D$ can be represented by a Petri Net since bounding the set of values allows us to represent each thread with a finite-state machine. For a fixed set of priorities $P$, the register automata $A(\Gamma)$ can be transformed to finite-state automata since the number of possible valuations of the registers is bounded.}

{\color{red} The set of executions $\mathcal{C}_D$ can be represented by a Vector Addition Systems with States (VASS), since both values and priorities are finite, which implies that each thread and register automata can be transformed into finite-state automata. To obtain this transformation, the states of the VASS represent the global variables of $\mathcal{C}_D$, while each counter of the VASS then represents the number of threads which are at a particular control location within a method, with a certain valuation of the local variables. In this way we transform linearizability problem of priority queue into state-reachability problem of VASS, which is EXPSPACE-complete. When we consider only finite number of threads, the value of counters are bounded and then this problem is PSPACE \cite{DBLP:conf/ac/Esparza96}.

On the other hand, we can mimic transitions of VASS by a priority queue implementation. To obtain this, we keep $\textit{rm}$ unchanged, while in each $\textit{put}$ method, we first put a value into priority queue, and then either simulate one step of transitions or record one increase/decrease of a counter, similarly as that in \cite{conf/esop/BouajjaniEEH13}. In this way we transform the state-reachability problem of VASS into linearizability problem of priority queue. Based on above discussion, we have the following complexity result. The detailed proof can be found in Appendix \ref{subsec:appendix proof and definition in subsection decidability result}.}

>>>>

\begin{lemma}
There exists a $\Gamma$-complete automaton for each $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMax}$ $\mathsf{Priority}$, $\mathsf{MatchedMaxPriority}\}$.
\end{lemma}

When defining $\Gamma$-complete automata, we assume that every implementation $\mathcal{I}$ behaves correctly, i.e., as a FIFO queue, when only values with the same priority are observed. More precisely, we assume that for every execution $e\in\mathcal{I}$ and every priority $p\in\mathbb{P}$, the projection of $e$ to values with priority $p$ is linearizable (w.r.t. $\seqPQ$). This property can be checked separately using register automata obtained from the finite automata in~\cite{DBLP:conf/icalp/BouajjaniEEH15} for FIFO queue (see Appendix~\ref{sec:appendix lemma and register automata for FIFO of single-priority executions} for more details). Note that this assumption excludes some obvious violations, such as a $\textit{rm}(a)$ operation happens before a $\textit{put}(a,p)$ operation, for some $p$.

@@

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Let $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$ and $e$ a data-differentiated execution. Then, $e$ is $\Gamma$-linearizable iff $e\vert_{\preceq p}$ is $\Gamma$-linearizable for some maximal priority $p$ in $e$.
\end{restatable}
\begin {proof} (Sketch)
%{\color {blue}We prove an equivalent property: $e$ is $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is $\Gamma$-linearizable for some maximal priority $p$ in $e$.} %We consider the case of $\mathsf{MatchedMaxPriority}$ and the case of $\mathsf{UnmatchedMaxPriority}$ is similar.
To prove the $\textit{only if}$ direction, let $e$ be a data-differentiated execution linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$. Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}$ $(s,x)$ imposes no restriction on the operations in $u$, $v$ and $w$ with priorities incomparable to $p$, erasing all these operations results in a sequential execution which still satisfies this property. Similarly, for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that is not $\mathsf{MatchedMaxPriority}^{>}$-linearizable. We represent each operation as a time interval whose left, resp., right, bound corresponds to the call, resp., return action. Operations adding and removing the same value are aligned vertically.}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

@@

The execution in \figurename~\ref{fig:introduce gap for EPQ1Lar} is not $\mathsf{MatchedMaxPriority}^>$-linearizable because the left-right constraint of the maximal priority value $b$ contains a cycle: $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$. The presence of such a cycle is equivalent to the execution not being $\mathsf{MatchedMaxPriority}^>$-linearizable (see Appendix~\ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}), as indicated by the following lemma:

\begin{restatable}{lemma}{LinEqualsConstraintforEPQOneLar}
\label{lemma:Lin Equals Constraint for EPQ1Lar}
Given a data-differentiated execution $e$ where $\mathsf{Has\text{-}MatchedMaxPriority}(e)$ holds, let $p$ be its maximal priority and $\textit{put}(x,p),\textit{rm}(x)$ are only operations of priority $p$ in $e$. Let $G$ be the graph representing the left-right constraint of $x$. $e$ is $\mathsf{MatchedMaxPriority}$-linearizable, if and only if $G$ has no cycle going through $x$.
\end{restatable}

@@

%TODO WHY IT IS A SET OF AUTOMATA ? SAY WHAT ARE THE CASES THAT NEED TO BE COVERED. SAY WHAT IS THE CASE WE GIVE IN THE FIGURE. USE $\top$ AS MENTIONED AT THE BEGINNING OF THE SECTION INSTEAD OF 	$c$. USE BIG CAPS FOR SETS OF TRANSITION LABELS.
To define a $\mathsf{MatchedMaxPriority}^>$-complete automaton we need to consider several cases depending on the order between the call/return actions of the $\textit{put}$/$\textit{rm}$ operations that add and respectively, remove the value $b$. For example, the case where the put happens-before the remove (as in \figurename~\ref{fig:introduce gap for EPQ1Lar}) is pictured in \figurename~\ref{fig:automata APQ1Lar-1 in paper}. This automaton captures the three possible ways of ordering the first action $\textit{ret}(\textit{put},a,\_)$ w.r.t. the actions with value $b$, which are pictured in \figurename~\ref{fig:executions APQ1Lar-1 in paper} (a) (this action cannot occur after $\textit{call}(\textit{rm},b,\_)$ since $b$ must be covered by the $a$-s). The paths corresponding to these three possible orders are: $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_7$, $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_{10}$ and $q_1 \rightarrow q_9 \rightarrow q_{10} \ldots \rightarrow q_7$. In \figurename~\ref{fig:executions APQ1Lar-1 in paper}, we show all the four orders of the call/return actions of adding and removing $b$, and also possible orders of the first $\textit{ret}(\textit{put},a,\_)$ w.r.t the actions with value $b$. In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, three register automata is constructed according to the cases of \figurename~\ref{fig:executions APQ1Lar-1 in paper} (b), (c) and (d), respectively.

@@

\begin{figure}[htbp]
  \centering
  %\includegraphics[width=1 \textwidth]{figures/PIC_HIS_PQ1Lar-pprr-in-paper.pdf}
  \includegraphics[width=1 \textwidth]{figures/PIC_HIS_PQ1Lar-fouCase.pdf}
%\vspace{-10pt}
  %\caption{Three possible positions of the first $\textit{call}(\textit{put},a,\_)$ in captured execution}
  \caption{Four cases of ordering actions with value $b$, and possible orders of the first $\textit{ret}(\textit{put},a,\_)$ w.r.t the actions with value $b$}
  \label{fig:executions APQ1Lar-1 in paper}
\end{figure}

@@

To recognize the scenarios in \figurename~\ref{fig:introduce pb order}, we introduce an order $<_{\textit{pb}}$ between values which intuitively, can be thought of as ``a value $a$ is put before another value $b$''. %Let us introduce put-before order $<_{\textit{pb}}$ to formally state that ``an value is putted before another value''.
Thus, given a data-differentiated execution $e$ and two values $a$ and $b$ of maximal priority, $a <_{\textit{pb}} b$ if one of the following holds: (1) $\textit{put}(a,\_) <_{\textit{hb}} \textit{put}(b,\_)$, (2) $\textit{rm}(a) <_{\textit{hb}} \textit{rm}(b)$, or (3) $\textit{rm}(a) <_{\textit{hb}} \textit{put}(b,\_)$. Sometimes we use $a <_{\textit{pb}}^A b$, $a <_{\textit{pb}}^B b$ and $a <_{\textit{pb}}^C b$ to explicitly distinguish between these three cases. Let $<_{\textit{pb}}^*$ be the transitive closure of $<_{\textit{pb}}$.
%Intuitively, $a <_{\textit{pb}}^* b$ means that we can infer that $a$ is putted before $b$ with the help of several intermediate values.

Then, to model the time intervals in which a remove operation like $\textit{rm}(b)$ in \figurename~\ref{fig:introduce pb order}, can be linearized (outside of intervals of smaller priority values) we introduce the notion of gap-point. Here we assume that the index of actions of an execution starts from $0$.

@@

\begin{restatable}{lemma}{EPQOneEqualAsPBandGP}
\label{lemma:EPQ1Equal as pb order and gap-point}
Let $e$ be a data-differentiated execution which contains only one maximal priority $p$ such that $\mathsf{Has\text{-}MatchedMaxPriority}(e)$ holds.
%Given a data-differentiated $p$-execution $e$ where $\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Condition}(e)$ holds.
Then, $e$ is not linearizable w.r.t $\mathsf{MatchedMaxPriority}^{=}$ iff $e$ contains two values $x$ and $y$ of maximal priority $p$ such that $y <_{\textit{pb}}^* x$, and the rightmost gap-point of $x$ is strictly smaller than the index of $\textit{call}(\textit{put},y,p)$ or $\textit{call}(\textit{rm},y)$.
\end{restatable}

@@

\smallskip
To characterize violations to linearizability w.r.t. $\mathsf{MatchedMaxPriority}^{=}$, one has to consider all the possible orders between call/return actions of the operations on values $a$, $b$, and $a_i$ in Lemma \ref{lemma:ob order has bounded length}, and the right-most gap point of $b$. Excluding the inconsistent cases, we are left with 5 possible orders that are shown in \figurename~\ref{fig:five enumerations}, where $o$ denotes the rightmost gap-point of $b$.
For each case, we define an automaton recognizing the induced set of violations. For instance, the register automata for the case of \figurename~\ref{fig:five enumerations}(a) is shown in \figurename~\ref{fig:an enumeration and its witness automaton}. In this case, the conditions in Lemma~\ref{lemma:EPQ1Equal as pb order and gap-point} are equivalent to the fact that intuitively, the time interval from $\textit{call}(\textit{rm},a)$ to $\textit{ret}(\textit{rm},b)$ is covered by lower priority values (and thus, there is no gap-point of $b$ which occurs after $\textit{call}(\textit{rm},a)$). Using again the data-independence property, these lower priority values can all be renamed to a fixed value $d$, and the other values to a fixed value $\top$.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8 \textwidth]{figures/PIC-HIS-FiveEnumerations.pdf}
%\vspace{-10pt}
  \caption{Five cases that need to be considered.}
  \label{fig:five enumerations}
\end{figure}

@@

We describe a class $\mathcal{C}$ of data-independent implementations for which linearizability w.r.t. $\seqPQ$ is decidable. The implementations in $\mathcal{C}$ allow an unbounded number of values but a bounded number of priorities. Each method manipulates a finite number of local variables which store Boolean values, or data values from $\mathbb{D}$. Methods communicate through a finite number of shared variables that also store Boolean values, or data values from $\mathbb{D}$. Data values may be assigned, but never used in program predicates (e.g., in the conditions of if-then-else and while statements) so as to ensure data independence. This class captures typical implementations, or finite-state abstractions thereof, e.g., obtained via predicate abstraction. Since the $\Gamma$-complete automata $A(\Gamma)$ uses a fixed set $D=\{a,b,a_1,d,e,\top\}$ of values, we have that $\mathcal{C}\cap A(\Gamma)\neq\emptyset$ for some $\Gamma$ iff $\mathcal{C}_D\cap A(\Gamma)\neq\emptyset$ where $\mathcal{C}_D$ is the subset of $\mathcal{C}$ that uses only values in $D$.

The set of executions $\mathcal{C}_D$ can be represented by a Vector Addition Systems with States (VASS), since both values and priorities are finite, which implies that each thread and register automata can be transformed into finite-state automata. To obtain this transformation, the states of the VASS represent the global variables of $\mathcal{C}_D$, while each counter of the VASS then represents the number of threads which are at a particular control location within a method, with a certain valuation of the local variables. In this way we transform linearizability problem of priority queue into state-reachability problem of VASS, which is EXPSPACE-complete. When we consider only finite number of threads, the value of counters are bounded and then this problem is PSPACE \cite{DBLP:conf/ac/Esparza96}.

On the other hand, we can mimic transitions of VASS by a priority queue implementation. To obtain this, we keep $\textit{rm}$ unchanged, while in each $\textit{put}$ method, we first put a value into priority queue, and then either simulate one step of transitions or record one increase/decrease of a counter, similarly as that in \cite{conf/esop/BouajjaniEEH13}. In this way we transform the state-reachability problem of VASS into linearizability problem of priority queue. Based on above discussion, we have the following complexity result. The detailed proof can be found in Appendix \ref{subsec:appendix proof and definition in subsection decidability result}.

### 2017-07-01 priority-queues-arxiv.tex
modify

We define $\seqPQ$, the set of sequential priority queue executions, semantically via labelled transition system ($LTS$). An LTS is a tuple $A=(Q,\Sigma,\rightarrow,q_0)$, where $Q$ is a set of states, $\Sigma$ is an alphabet of transition labels, $\rightarrow\subseteq Q\times\Sigma\times Q$ is a transition relation and $q_0$ is the initial state. We model priority queue as an LTS $\textit{PQ}$ defined as follows: {\color {blue}each state of $\textit{PQ}$}{\color {red}each state of the LTS $\textit{PQ}$} is a mapping associating priorities in $\mathbb{P}$ with sequences of values in $\mathbb{D}$ representing a snapshot of the priority queue (for each priority, the values are ordered as they were inserted); the transition labels are $\textit{put}(a,p)$ and $\textit{rm}(a)$ operations; Each transition modifies the state as expected. For example, $q_1 \xrightarrow{\textit{rm}(\textit{empty})} q_2$ if $q_1 = q_2$, and $q_1$ and $q_2$ map each priority to the empty sequence $\epsilon$. Then, {\color{blue}$\seqPQ$ is the set of words accepted by $\textit{PQ}$}{\color {red}$\seqPQ$ is the set of traces of $\textit{PQ}$}. The detailed definition of $\textit{PQ}$ can be found in Appendix \ref{sec:appendix definition of seqPQ and proof of Lemma EQP rules and semantics}.

@@

Given a (concurrent) execution $e$ and a sequential execution $s$, we say that $e$ is linearizable w.r.t $s$, denoted $e \sqsubseteq s$, if there is a bijection $f: O_1 \rightarrow O_2$, where $O_1$ and $O_2$ are the set of operations of $e$ and $s$, respectively, such that (1) {\color{blue}$o$ and $f(o)$ have the same call and return actions}{\color {red}$o$ and $f(o)$ is the same operation}\footnote{An $m(a)$-operation in an execution $e$ is an operation identifier $o$ s.t. $e$ contains the actions $\textit{call}_o(m,a)$, $\textit{ret}_o(m,a)$.}, and (2) if $o_1 <_{\textit{hb}} o_2$, then $f(o_1) <_{\textit{hb}} f(o_2)$.
A (concurrent) execution $e$ is linearizable w.r.t a set $S$ of sequential executions, denoted $e \sqsubseteq S$, if there exists $s \in S$ such that $e \sqsubseteq s$. A set of concurrent executions $E$ is linearizable w.r.t $S$, denoted $E \sqsubseteq S$, if $e \sqsubseteq S$ for all $e \in E$.

>>>>

We define $\seqPQ$, the set of sequential priority queue executions, semantically via labelled transition system ($LTS$). An LTS is a tuple $A=(Q,\Sigma,\rightarrow,q_0)$, where $Q$ is a set of states, $\Sigma$ is an alphabet of transition labels, $\rightarrow\subseteq Q\times\Sigma\times Q$ is a transition relation and $q_0$ is the initial state. We model priority queue as an LTS $\textit{PQ}$ defined as follows: each state of the LTS $\textit{PQ}$ is a mapping associating priorities in $\mathbb{P}$ with sequences of values in $\mathbb{D}$ representing a snapshot of the priority queue (for each priority, the values are ordered as they were inserted); the transition labels are $\textit{put}(a,p)$ and $\textit{rm}(a)$ operations; Each transition modifies the state as expected. For example, $q_1 \xrightarrow{\textit{rm}(\textit{empty})} q_2$ if $q_1 = q_2$, and $q_1$ and $q_2$ map each priority to the empty sequence $\epsilon$. Then, $\seqPQ$ is the set of traces of $\textit{PQ}$. The detailed definition of $\textit{PQ}$ can be found in Appendix \ref{sec:appendix definition of seqPQ and proof of Lemma EQP rules and semantics}.

@@

Given a (concurrent) execution $e$ and a sequential execution $s$, we say that $e$ is linearizable w.r.t $s$, denoted $e \sqsubseteq s$, if there is a bijection $f: O_1 \rightarrow O_2$, where $O_1$ and $O_2$ are the set of operations of $e$ and $s$, respectively, such that (1) $o$ and $f(o)$ is the same operation\footnote{An $m(a)$-operation in an execution $e$ is an operation identifier $o$ s.t. $e$ contains the actions $\textit{call}_o(m,a)$, $\textit{ret}_o(m,a)$.}, and (2) if $o_1 <_{\textit{hb}} o_2$, then $f(o_1) <_{\textit{hb}} f(o_2)$.
A (concurrent) execution $e$ is linearizable w.r.t a set $S$ of sequential executions, denoted $e \sqsubseteq S$, if there exists $s \in S$ such that $e \sqsubseteq s$. A set of concurrent executions $E$ is linearizable w.r.t $S$, denoted $E \sqsubseteq S$, if $e \sqsubseteq S$ for all $e \in E$.
